<<<<<<< HEAD
---
title: "Biostatistics project - Group 25"
author: "Emma Van Kerckhove, Kylian Baeyens, Robbe Claeys, Tom Nissens"
date: "`r Sys.Date()`"
output: html_document
---

# BASTA strawberry trial 2022
The biostatistics course project is based on the BASTA research (Biochar’s Added value) organised by ILVO, UHasselt and pcfruit.

Biochar is the residual product formed after pyrolysis of biomass and has currently no economic value. In this project researchers are interested to see if (1) Biochar can partly replace peat and (2) Biochar can be used to promote plant development and/or plant disease resistance of strawberry.

---

**Variables:**

- the type of growing medium (peat, peat_biochar, peatfree, peatfree_biochar)
- the addition of biochar as a vaccine (vaccine 1-3)
- the addition of *Trichoderma* (*Trichoderma* 1-3)

There were a total of 28 growing blends (4 x 3 x 3 + 4 controls). For each blend three plant trays were filled, each containing five strawberry plants, and replicated four times. The trays were divided over two greenhouses (greenhouse 1-2).

**Measurements:**

- root rot score (30, 50, and 80 days after infection)
- total number of fruits over 15 plants
- number of fruits infected with *Botrytis* (after harvest, one week storage at 6°C, and three days storage at 20°C)
- weight of strawberries per plant in kg (class 1, class 2, curved, and total strawberries)
- fresh biomass per plant
- dry biomass per plant

**Research question:**  
On average, is there an effect of growing medium and/or the addition of Trichoderma on the weight of the curved strawberries. (data set 2)

---

# Data Preparation
## Load the libraries

```{r results = 'hide', warning = FALSE}
#library(ggplot2)
#library(dplyr)
library(tidyverse)
library(car)
library(multcomp)
library(SummarizedExperiment)
library(edgeR)
library(pwr)
```


## Read the data

```{r}
data <- read.delim("data/BASTA_Dataset2_Trichoderma.txt", header = TRUE, stringsAsFactors = FALSE, quote = "", sep = "\t")
head(data)
```

# Research Investigation
## Variables

**Response variable:** 
The weight of the curved strawberries: `r colnames(data[19])`

**Explanatory variable(s):**  
- Growing medium (peat/non-peat and biochar/ no biochar): `r colnames(data[5])`
- Addition of *Trichoderma*: `r colnames(data[9])`
- Greenhouse: `r colnames(data[3])`

**Observational unit:** 
The strawberry plant.

**Experimental unit:** 
Three trays of 5 plants.

## Hypotheses

H01: On average, the growing medium has no effect on the weight of the curved strawberries.  

H02: On average, the addition of *Trichoderma* has no effect on the weight of the curved strawberries.  

H03: The effect of *Trichoderma* addition on the weight of the curved strawberries does not differ for the different growth media.  


Ha1: On average, the growing medium has an effect on the weight of the curved strawberries.  

Ha2: On average, the addition of *Trichoderma* has an effect on the weight of the curved strawberries.  

Ha3: The effect of *Trichoderma* addition on the weight of the curved strawberries differs for the different growth media.  

# Data Exploration
First we filter the data, generate an appropriate format, and provide some basic statistics.

```{r}
df <- data.frame(data[,c(1, 4:5, 9, 19)])
df$Greenhouse <- factor(df$Greenhouse)
df$Trichoderma <- factor(df$Trichoderma)
df$Substrate <- factor(df$Substrate)
df$SampleName <- factor(df$SampleName)

head(df)
summary(df)
```

Next, some basic plots are generated to provide an initial intuitive view on the effects of the explanatory variables.

```{r}
# plots effect of substrate choice
df %>% 
  ggplot(aes(x = Substrate, y = Curved_strawberries, fill = Substrate)) +
  geom_boxplot() + 
  labs(
    title = "Effect of substrate choice on the weight of curved strawberries", 
    y = "Curved strawberry weight per plant (kg)") + 
  geom_jitter(width = 0.2, alpha = 0.3) +
  theme_bw()

# plots effect of Trichoderma addition
df %>% 
  ggplot(aes(x = Trichoderma, y = Curved_strawberries, fill = Trichoderma)) +
  geom_boxplot() + 
  labs(
    title = "Effect of Trichoderma addition on the weight of curved strawberries", 
    y = "Curved strawberry weight per plant (kg)") + 
  geom_jitter(width = 0.2, alpha = 0.3) +
  theme_bw()

# plots effect of greenhouse choice
df %>% 
  ggplot(aes(x = Greenhouse, y = Curved_strawberries, fill = Greenhouse)) +
  geom_boxplot() + 
  labs(
    title = "Effect of greenhouse choice on the weight of curved strawberries", 
    y = "Curved strawberry weight per plant (kg)") + 
  geom_jitter(width = 0.2, alpha = 0.3) +
  theme_bw()

# bar chart to check for possible interactivity between variables
df %>%
  filter(Trichoderma %in% c("None","Trich1","Trich2","Trich3")) %>%
  ggplot(aes(x = Substrate, y = Curved_strawberries, fill = Trichoderma)) +
  geom_bar(stat = "identity") + 
       labs(
         title = "Interactive effect of the substrate and trichoderma addition\non the weight of curved strawberries",
         x = "Type of substrate",
         y = "Weight of curved strawberries") +
  theme_bw()
```

At first sight We can see that substrate choice (and to a lesser extent *Trichoderma* addition) on average seems to have an effect on the weight of curved strawberries. The choice of greenhouse, however, does not seem to have an effect but a slightly higher variance can be observed for greenhouse 1. There seems to be no equality of variances (heteroscedasticity) between the 16 different combinations upon viewing the height of the barcharts from any of the frames generated.

To investigate further, a plot was generated that distinguishes between the substrate choice & *Trichoderma* addition effects. This leads to believe that the effects of both explanatory variables are linked.

```{r}
# plots effect of substrate choice & Trichoderma addition
df %>% 
  filter(Trichoderma %in% c("None","Trich1","Trich2","Trich3")) %>%
  ggplot(aes(x = Substrate, y = Curved_strawberries, fill = Trichoderma)) +
  geom_boxplot() + 
    labs(
    title = "Effect of substrate choice & Trichoderma addition\non the weight of curved strawberries", 
    y = "Curved strawberry weight per plant (kg)") + 
  geom_jitter(width = 0.2, alpha = 0.3) +
  theme_bw()
```

Due to the limited number of observations and the large number of variables, trends in these plots should not be taken for granted and further statistical investigation will be performed.

# Model Selection & Statisticial Inference

## Assumption checking
To check normality, histograms & QQ-plots are generated.

```{r}
df %>% 
  ggplot(aes(x = Curved_strawberries)) +
  geom_histogram(fill = "#D95C68", binwidth = 0.001) +
      labs(
    title = "Normality confirmation of curved strawberry weight measures: Histogram", 
    x = "Curved strawberry weight per plant (kg)") + 
  theme_bw()

df %>% 
  ggplot(aes(sample = Curved_strawberries)) +
  stat_qq() +
  stat_qq_line(col = "#D95C68") +
  labs(
    title = "Normality confirmation of curved strawberry weight measures: QQ-plot", 
    x = "Theoretical Quantiles",
    y = "Sample Quantities") + 
  theme_bw()
```

The data are not normally distributed, has a lower bound of zero, and a tail to the right can be observed. A square root-transformation might help to make the data more normally distributed. The assumptions are re-evaluated.

```{r warning=FALSE}
df_t <- df
df_t$Curved_strawberries <- ((df_t$Curved_strawberries)^0.5)

df_t %>% 
  ggplot(aes(x = Curved_strawberries)) +
  geom_histogram(fill = "#D95C68") +
      labs(
    title = "Normality confirmation of curved strawberry weight measures\nafter square root-transormation: Histogram", 
    x = "Curved strawberry weight per plant (kg)") + 
  theme_bw()

df_t %>% 
  ggplot(aes(sample = Curved_strawberries)) +
  stat_qq() +
  stat_qq_line(col = "#D95C68") +
  labs(
    title = "Normality confirmation of curved strawberry weight measures\nafter square root-transformation: QQ-plot", 
    x = "Theoretical Quantiles",
    y = "Sample Quantities") + 
  theme_bw()
```

A square root transformation has a beneficial effect on the normality of the data, the departures from normality are also symmetric. Back-transformation will have to be performed during model interpretation. *It was noticed however, that there is an observation where the weight of curved strawberries is zero. This value doesn't hinder further analysis and is still taken into account.

Although homoscedasticity seems to be met (based on the previously shown box plots), an additional sanity check is performed with Levene’s test (which is robust against normality violations). If the resulting p-value is greater than 0.05, the null-hypothesis cannot be rejected - meaning that sample variance is likely to have occurred based on random sampling from a population with equal variances.

```{r warning=FALSE}
leveneTest(Curved_strawberries ~ interaction(Substrate, Trichoderma), data = df_t)
```

## Model fitting

A model is generated with substrate-choice, *Trichoderma*-addition, and an interaction variable between both as explanatory variables.

```{r}
# set peatfree as intercept (peatfree is control soil)
df_t <- within(df_t, Substrate <- relevel(Substrate, ref = "Peatfree"))

fit1 = lm(Curved_strawberries ~ Substrate + Trichoderma + Substrate:Trichoderma, data = df_t)
summary(fit1)
plot(fit1)
```


The assumptions and testing plots show good results but no 5% significance is obtained for the overall model (p-value: 0.07216). Analysis of variance (ANOVA) is performed to check which variables best explain the variance present in the data.

```{r}
anova(fit1)
```

When performing anova, only a significant result can be obtained for substrate as explanatory variable.

Additionaly, the variance inflation factor (VIF) for the fit1 can indicate whether there is multicollinearity between two or more variables.

```{r, warning=FALSE}
vif(fit1, type = "predictor")

```

The GVIF values (generalized VIF) indicate there is no correlation between the predictor variables in the model, therefore, all variables of fit1 are kept in the final model.

Linear model fit with only substrate as explanatory variable:

```{r}
fit2 <- lm(Curved_strawberries ~ Substrate, data = df_t)
summary(fit2)
plot(fit2)
```

The assumptions and testing plots show normality.
An overall p-value of 0.0442 for the model is obtained, indicating that substrate is indeed the primary driver of curved strawberry weight differences. 

### Blocking

The greenhouse choice could be responsible for technical variation present in the data. As the greenhouses can have a slightly different temperature, humidity, ... Therefore, the greenhouse variable is incorporated in the model. (blocking)

```{r}
fit3 = lm(Curved_strawberries ~ Substrate + Trichoderma + Greenhouse + Substrate:Trichoderma, data = df_t)
summary(fit3)
plot(fit3)
```

The model produces less significant results when incorporating the greenhouse choice as a parameter,(p-value: 0.09714 vs. 0.07216) but it is kept in the model in order to account for possible technical variation.

Further statistical testing and pairwise comparison is performed, although the model isn't significant.

```{r, warning=FALSE}
vif(fit3, type = "predictor")
```

The GVIF values are close to 1, indicating that there is no correlation between any explanatory variables after adding the greenhouse variable.

```{r}
anova_fit3 <- aov(Curved_strawberries ~ Substrate + Trichoderma + Greenhouse + Substrate:Trichoderma , data = df_t)
summary(anova_fit3)
```

Performing an ANOVA test, there can be seen that there is only a significant difference between the different substrates. A Tukey post-hoc analysis is performed to mutual compare the different substrates. 

```{r}
# this is how it is performed in practical session 1 exercise 2 (mss weglaten???)
Tukey_fit3 <- TukeyHSD(anova_fit3, which = "Substrate")
Tukey_fit3

```

It can be concluded that only a significant difference in weight of curved strawberries can be found when comparing peatfree substrate with biochar and substrate with peat and biochar (p-value = 0,021). 


```{r}
#Interpretation on original scale
sqroot <- summary(fit3)
(sqroot$coefficients)^2
(confint(fit3))^2
# Ik weet niet hoe de waarden van de Tukey terug te transformeren. Mss gaat dit niet?
(Tukey_fit3$diff)^2
```

The lack of significant findings could be due to a lack of power. A power analysis is performed
```{r}
pwr.f2.test(u=4,v=64-4-1,f2=0.1384/(1-0.1384),sig.level = 0.05, power = NULL)
```

The power of our lineair model is 69%. This indicates that there is a 69% probability of not making a type 2 error. This could be improved by repeating the experiment with more samples. Next, we determine the amount of samples should be taken to improve our model to 80%.

```{r}
pwr.f2.test(u=4,v=NULL, f2=0.1384/(1-0.1384), sig.level=0.05, power=0.8)
```

To improve the power of our model to 80%, at least 80 samples would have to be taken.

# Conclusion
From our lineair model ()



The project needs to be handed in by Friday 23/12. Hand in an HTML file containing all your code + conclusions. The file should be saved as ProjectBiostatistics_Member1_Member2_Member3_Member4.
